Analyzing Income Inequality Trends in the Canada: Impacts of the Great Recession (2007-2009) and COVID-19 Pandemic (2020-2023)

### 2024-12-02 paper/paper.qmd
Update paper.qmd

---
title: "Analyzing Income Inequality Trends in the Canada: Impacts of the Great Recession (2007-2009) and COVID-19 Pandemic (2020-2023)"
subtitle: "Insights from 2002 to 2023"
author: Dong Jun Yoon
thanks: "Code and data are available at: https://github.com/DJY1231/Exploring_Income_Inequality_in_Canada_During_Major_Recessions.git"
date: today
date-format: long
abstract: "This paper examines income inequality in Canada from 2002 to 2023 using data from the World Inequality Database (WID) and Statistics Canada. The analysis explores trends in income inequality and inflation, assessing whether these disparities have increased or decreased over time. It also compares the impacts of the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023) on income disparities. Through data analysis and regression modeling, this study investigates these trends in depth. Understanding these patterns is crucial for policymakers and the government to develop effective strategies aimed at reducing inequalities within the country. This report ultimately seeks to contribute to efforts promoting social and economic equity in Canada."
format: pdf
number-sections: true
bibliography: references.bib
---
# Introduction
Income inequality is a significant issue both in Canada and all around the world. Income inequality refers to the difference in how income is distributed among the population, highlighting significant disparities in financial well-being. Unlike wealth inequality, which considers the value of assets such as homes, stocks, and other holdings, income inequality specifically examines earnings from sources like wages, salaries, savings interest, stock dividends, rental income, and profits from the sale of assets. This distinction is essential for understanding the multifaceted nature of economic disparities. Gaining a deeper understanding of income inequality can equip policymakers and researchers with valuable insights to develop strategies aimed at addressing this pervasive problem.

This paper focuses on the dynamics of income inequality in Canada, especially focusing on the period from Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023). Using the dataset provided by World Inequality Database (WID), the paper will fully investigate income trends across the social class, top 1%, top 10% and bottom 50% of income inequality. Moreover, will be considering variables such as year, social class, and race to find out the correlation between these factors.

To efficiently investigate remaining income gap between individual firms and social classes who are living in Canada, this paper will focus on exploring datasets and multiple linear regression modeling to find out the full details (trends, patterns) about income inequality that is occurring in Canada.

The paper is structured as following: Section 2 introduces the data used for analysis and findings, including the variables of interest. Section 3 introduces the linear regression model to examine the correlation between time (year) and social class to investigate income inequality in Canada. Section 4 introduces the full analysis of the table which will include the summary. Section 5 may discuss about the findings that we investigated and also talk about the weaknesses of this paper and its next steps.

## Estimand

The goal of this paper is to find out the relationship between income inequality and time, measured in years in Canada, focusing on how it changes over the years affect income inequality. Furthermore, this paper investigates whether the Great Recession (2007–2009) or the COVID-19 pandemic (2020–2023) has caused more income inequality between each social classes. The measurement uses data from the World Inequality Database (WID) to graph the income inequality trends from 2002 to 2023. And also uses data from Statistic Canada to investigate the inflation rate within each year.

@@

# Data {#sec-data}
## Data source
The data is gathered from World Inequality Database (WID) which introduces the income share by each social class over time: 2002 to 2023 in Canada. The second data is gathered from Statistics Canada where the data illustrates the Consumer Price Index (CPI) to figure out the inflation rate within each year.

## Data mearsurement
The dataset provides information that offers insights into the income inequality trends in Canada. The y-axis represents the income share, measured in dollars (%) while the x-axis corresponds to the years from 2002 to 2023, representing the share of pre tax total income for each year.

@@

```{r results='hide', message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)

@@

## Model set-up
The multiple linear regression model is as follows: 
$$
\begin{align*}
\text{Share_of_total_income}(Y) &= \beta_0 + \beta_1 \cdot \text{Year} + \beta_2 \cdot \text{Social_Class} \\
&\quad + \beta_3 \cdot \text{Period} + \beta_4 \cdot (\text{Year} \times \text{Social_Class}) \\
&\quad + \beta_5 \cdot (\text{Social_Class} \times \text{Period}) + \varepsilon
\end{align*}
$$
### Model justification

We aim to model the relationship between the response variable Y, which represents the share of income for each social class, using the following components:

@@

## Comparing income inequality trends during two major recessions: the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023).

Based on Figure 2, which displays the trend of income distribution across different social classes in Canada from 2007 to 2009, the data indicates a downward trend in the income share for the top 1% and top 10% of earners during the period from 2007 to 2009, suggesting that these groups lost a portion of their total income share during the Great Recession. Specifically, the income share of the top 1% decreased by 2.35 percentage points, from 14.21% to 11.86% of the total income pie. Similarly, the top 10% experienced a decrease of 2.5 percentage points, from 38.38% to 35.88%. During this period, many businesses and small retailers faced significant challenges, forcing them to reduce spending, cut investments, and, in some cases, declare bankruptcy due to low liquidity and limited access to credit. This resulted in widespread job losses, reduced incomes, and an overall economic downturn, ultimately contributing to a contraction in consumer spending and slowing economic recovery.

@@

## How Consumer Price Index (CPI) rates may contribute to increasing or decreasing income inequality.

Figure 4 illustrates the trend of the Consumer Price Index (CPI) in Canada from 2002 to 2023. The CPI shows a consistent upward trend over this period, indicating that consumer prices have steadily increased. This trend suggests that rising consumer prices are associated with an increase in the inflation rate. In other words, as the inflation rate rises, purchasing power—defined as the ability to buy products and services—decreases. We measure inflation by the CPI annual growth rate, which directly reflects the increase in the cost of essential goods and services.

@@

## Weaknesses and Limitations

While the analysis provides valuable insights, several limitations remain. First, it relies on aggregated data that may mask important regional or demographic variations in income inequality trends. For example, certain provinces or industries in Canada may have experienced different degrees of impact during these recessions. Second, the analysis focuses on income share percentages but does not focuses into the absolute changes in wealth or income levels, which could provide a more comprehensive picture of inequality. Additionally, while the regression modeling captures general trends, it may not fully account for other confounding factors, such as shifts in taxation policies, labor market dynamics, or global economic conditions, which also influence income distribution. Lastly, the analysis does not explore the long-term implications of these recessions, such as their effects on intergenerational mobility or wealth accumulation.

@@

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

@@

Data - wid - world inequality database. WID. (n.d.). https://wid.world/data/

>>>>

---
title: "Analyzing Income Inequality Trends in the Canada: Impacts of the Great Recession (2007-2009) and COVID-19 Pandemic (2020-2023)"
subtitle: "Insights from 2002 to 2023"
author: Dong Jun Yoon
thanks: "Code and data are available at: https://github.com/DJY1231/Exploring_Income_Inequality_in_Canada_During_Major_Recessions.git"
date: today
date-format: long
abstract: "This paper examines income inequality in Canada from 2002 to 2023 using data from the World Inequality Database (WID) and Statistics Canada. The analysis explores trends in income inequality and inflation, assessing whether these disparities have increased or decreased over time. It also compares the impacts of the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023) on income disparities. Through data analysis and regression modeling, this study investigates these trends in depth. Understanding these patterns is crucial for policymakers and the government to develop effective strategies aimed at reducing inequalities within the country. This report ultimately seeks to contribute to efforts promoting social and economic equity in Canada."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
# Introduction
Income inequality is a significant issue both in Canada and all around the world. Income inequality refers to the difference in how income is distributed among the population, highlighting significant disparities in financial well-being. Unlike wealth inequality, which considers the value of assets such as homes, stocks, and other holdings, income inequality specifically examines earnings from sources like wages, salaries, savings interest, stock dividends, rental income, and profits from the sale of assets. This distinction is essential for understanding the multifaceted nature of economic disparities. Gaining a deeper understanding of income inequality can equip policymakers and researchers with valuable insights to develop strategies aimed at addressing this pervasive problem.

This paper focuses on the dynamics of income inequality in Canada, especially focusing on the period from Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023). Using the dataset provided by World Inequality Database (WID), the paper will fully investigate income trends across the social class, top 1%, top 10% and bottom 50% of income inequality. Moreover, will be considering variables such as year and CPI to find out the correlation between income inequality and inflation.

To efficiently investigate remaining income inequality between individual firms and social classes who are living in Canada, this paper will focus on exploring datasets and multiple linear regression modeling to find out the full details (trends, patterns) about income inequality that is occurring in Canada.

The paper is structured as following: Section 2 introduces the data used for analysis and findings, including the variables of interest. Section 3 introduces the linear regression model to examine the correlation between time (year) and social class to investigate income inequality in Canada. Section 4 introduces the full analysis of the table which will include the summary. Section 5 may discuss about the findings that we investigated and also talk about the weaknesses of this paper and its next steps. Section 6 will discuss about the focus on an aspect of surveys, sampling or observational data, related to the paper.

## Estimand
The goal of this paper is to explore the relationship between income inequality and time, measured in years, in Canada, with a focus on how changes over the years affect income inequality. Additionally, this paper examines whether the Great Recession (2007–2009) or the COVID-19 pandemic (2020–2023) caused greater income inequality among different social classes. The analysis uses data from the World Inequality Database (WID) to graph income inequality trends from 2002 to 2023, supplemented by data from Statistics Canada to examine the inflation rate for each year.

@@

# Data {#sec-data}
## Data source
The data is gathered from World Inequality Database (WID) which introduces the pre tax income share by each social class over time: 2002 to 2023 in Canada. The second data is gathered from Statistics Canada where the data illustrates the Consumer Price Index (CPI) to figure out the inflation rate within each year.

## Data mearsurement
The WID dataset provides information that offers insights into the income inequality trends in Canada. The y-axis represents the income share, measured in dollars (%) while the x-axis corresponds to the years from 2002 to 2023, representing the share of pre tax total income for each year. For the Statistics Canada dataset, it provides the information of yearly Consumer Price Index (CPI). The y-axis represents the CPI, measured in index base period while the x-axis corresponds to the years from 2002 to 2023, representing the CPI for each year.

@@

```{r}
# Load necessary libraries
library(ggplot2)

@@

## Model set-up
The multiple linear regression model is as follows: 
$$
\begin{align*}
\text{Share_of_total_income}(Y) &= \beta_0 + \beta_1 \cdot \text{Year} + \beta_2 \cdot \text{Social_Class} \\
&\quad + \beta_3 \cdot \text{Period} + \beta_4 \cdot (\text{Year} \times \text{Social_Class}) \\
&\quad + \beta_5 \cdot (\text{Social_Class} \times \text{Period}) + \varepsilon
\end{align*}
$$
### Model justification
We aim to model the relationship between the response variable Y, which represents the share of income for each social class, using the following components:

@@

## Comparing income inequality trends during two major recessions: the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023). 
Based on Figure 2, which displays the trend of income distribution across different social classes in Canada from 2007 to 2009, the data indicates a downward trend in the income share for the top 1% and top 10% of earners during the period from 2007 to 2009, suggesting that these groups lost a portion of their total income share during the Great Recession. Specifically, the income share of the top 1% decreased by 2.35 percentage points, from 14.21% to 11.86% of the total income pie. Similarly, the top 10% experienced a decrease of 2.5 percentage points, from 38.38% to 35.88%. During this period, many businesses and small retailers faced significant challenges, forcing them to reduce spending, cut investments, and, in some cases, declare bankruptcy due to low liquidity and limited access to credit. This resulted in widespread job losses, reduced incomes, and an overall economic downturn, ultimately contributing to a contraction in consumer spending and slowing economic recovery.

@@

## How Consumer Price Index (CPI) rates may contribute to increasing or decreasing income inequality.
Figure 4 illustrates the trend of the Consumer Price Index (CPI) in Canada from 2002 to 2023. The CPI shows a consistent upward trend over this period, indicating that consumer prices have steadily increased. This trend suggests that rising consumer prices are associated with an increase in the inflation rate. In other words, as the inflation rate rises, purchasing power—defined as the ability to buy products and services—decreases. We measure inflation by the CPI annual growth rate, which directly reflects the increase in the cost of essential goods and services.

@@

## Weaknesses and Limitations
While the analysis provides valuable insights, several limitations remain. First, it relies on aggregated data that may mask important regional or demographic variations in income inequality trends. For example, certain provinces or industries in Canada may have experienced different degrees of impact during these recessions. Second, the analysis focuses on income share percentages but does not focuses into the absolute changes in wealth or income levels, which could provide a more comprehensive picture of inequality. Additionally, while the regression modeling captures general trends, it may not fully account for other confounding factors, such as shifts in taxation policies, labor market dynamics, or global economic conditions, which also influence income distribution. Lastly, the analysis does not explore the long-term implications of these recessions, such as their effects on intergenerational mobility or wealth accumulation.

@@

# Appendix {-}
## Sketches
Data, graph, and analysis sketches can be found in the Github Repository.

## Additional data details
World Inequality Database (WID):
Data Description: The WID dataset provides detailed information on income distribution across social classes, including the top 1%, top 10%, and bottom 50% income earners in Canada. The dataset spans from 2002 to 2023, with annual observations. It includes pre-tax income shares expressed as a percentage of total income.

Geographic Coverage: The data is national in scope, representing Canada as a whole, without regional disaggregation.

Temporal Coverage: The dataset covers key economic events, such as the Great Recession (2007–2009) and the COVID-19 pandemic (2020–2023), allowing for trend analysis and comparative studies.

Statistics Canada (CPI Data):
Data Description: The Consumer Price Index (CPI) dataset measures the annual cost of a fixed basket of goods and services over time, serving as an indicator of inflation. The dataset includes values from 2002 to 2023.

Geographic Coverage: The CPI data represent a national average but include price trends from various provinces and urban areas.

Temporal Granularity: The CPI data are reported monthly, but for consistency with the WID dataset, annual averages were calculated.

## Posterior predictive check (PPC)

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Simulate data (replace this with your actual data and model)
set.seed(123)
data <- data.frame(
  Year = rep(2002:2023, each = 3),
  Social_Class = rep(c("Bottom 50%", "Top 10%", "Top 1%"), times = 22),
  Income_Share_Obs = c(runif(22, 15, 20), runif(22, 30, 35), runif(22, 10, 15))
)

# Fit a simple linear regression model (replace this with your actual model)
model <- lm(Income_Share_Obs ~ Year + Social_Class, data = data)

# Generate posterior predictive samples (for simplicity, simulate predictions here)
# Replace this with Bayesian sampling if applicable
data <- data %>%
  mutate(
    Income_Share_Pred = predict(model, newdata = data) + rnorm(nrow(data), mean = 0, sd = 0.5) # Adding noise
  )

# Posterior Predictive Check: Density Plot
ggplot(data, aes(x = Income_Share_Obs, fill = "Observed")) +
  geom_density(alpha = 0.5, color = NA) +
  geom_density(aes(x = Income_Share_Pred, fill = "Predicted"), alpha = 0.5, color = NA) +
  labs(
    title = "Posterior Predictive Check: Income Share Distribution",
    x = "Income Share (%)",
    y = "Density"
  ) +
  scale_fill_manual(name = "Legend", values = c("Observed" = "blue", "Predicted" = "gray")) +
  theme_minimal()

# Posterior Predictive Check: Time Series Comparison
ggplot(data, aes(x = Year)) +
  geom_line(aes(y = Income_Share_Obs, color = "Observed"), size = 1) +
  geom_line(aes(y = Income_Share_Pred, color = "Predicted"), size = 1, linetype = "dashed") +
  facet_wrap(~ Social_Class, scales = "free_y") +
  labs(
    title = "Posterior Predictive Check: Observed vs Predicted Trends",
    x = "Year",
    y = "Income Share (%)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Observed" = "blue", "Predicted" = "red")) +
  theme_minimal()

```
Figure 6: Posterior distribution for Income Inequality

Figure 6 visually displays how closely the predicted income shares generated by the model align with the observed income share. If the model is well-fitted, the predicted values will closely match the observed values in both distribution and trends.

@@

Data - wid - world inequality database. WID. (n.d.). https://wid.world/data/ 
Canada, S. (2024, January 16). Consumer price index, annual average, not seasonally adjusted. Consumer Price Index, annual average, not seasonally adjusted. https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1810000501
